Improving Item-Item Similarity Estimation during Collaborative Filtering

### 2014-03-26 paper/paper.tex
Extend results of rating-based method.

Fig. \ref{fig:RetingDist} shows density plots of rating pairs, for $N \geq 40$
and $N \geq 80$. These data, together with the similarity score histograms of Fig.
\ref{fig:Histograms}, suggest that the data for our two data sets are similar
and that large biases are not introduced by varying the number of common users,
$N$. At the same time, both rating distributions reveal significant correlations
between Rating1 and Rating2 as evidenced by the elongation in the $45^\circ$
direction. This is consistent with the data of Fig. \ref{fig:Histograms}, which
shows the mean $PearSim > 0$ for both sets. Further, these data indicate a
bias toward positive values among rating pairs compared to those of ratings in
general, which have an average value of zero by construction\footnote{See
discussion of bias subtraction in {\bf Data} section.}. This latter feature
indicates that items with large numbers of common users tend to be rated higher
by their common users than by their other users since we have already subtracted
the item's average rating thereby avoiding any simple popularity effects.

@@

Having chosen values for $\mu_s$, $\sigma_s$, $\mu_r$, and $\sigma_r$ we are now
able to estimate the similarity of item pairs. To do this, we must solve Eq.
(\ref{eqn:estimator}) for $\hat{y}$ given a pair of ratings. This can be easily
done in practice using a root finding algorithm\footnote{To solve Eq.
(\ref{eqn:estimator}) for $\hat{y}$ given ratings $r_A$ and $r_B$, we use the
{\tt brentq} root finding function of the widely available Python module {\tt
scipy.optimize}.}. The implied predictions of individual users can then be
aggregated to form a new estimate of the similarity of items $A$ and $B$ as
\begin{align}\label{eqn:userPredSim}
UserPredSim(A,B) = \frac{\sum_{u \in U_{AB}}{p_u\cdot\hat{y}(r_{u,A}, r_{u,B})}}
{\sum_{u \in U_{AB}}{p_u}},
\end{align}
where, as before, $U_{AB}$ is the set of common users for items $A$ and $B$ and
$p_u$ is the {\em user predictivity} defined in Eq. (\ref{eqn:predictivity}). We
leverage the training set to find $p_u$ for each user $u$. This is done by
recording the implied prediction $\hat{y}(r_{u,A},r_{u,B})$ for each common user
$u$ for each item pair $(A, B)$ in the training set and then computing each
user's average absolute error $\epsilon_u$ as per Eq. (\ref{eqn:userError}).
Finally, we apply supervised learning techniques to optimize the value of the
parameter $\lambda$ that appears in the definition of $p_u$. This is achieved by
minimizing the error in Eq. {\ref{eqn:userPredSim} made against the training
set.

>>>>

Figure \ref{fig:RatingDist} shows density plots of rating pairs, for $N \geq 40$
and $N \geq 80$. These data, together with the similarity score histograms of Fig.
\ref{fig:Histograms}, suggest that the data for our two data sets are similar
and that large biases are not introduced by varying the number of common users,
$N$. At the same time, both rating distributions reveal significant correlations
between Rating1 and Rating2 as evidenced by the elongation in the $45^\circ$
direction. This is consistent with the data of Fig. \ref{fig:Histograms}, which
shows the mean $PearSim > 0$ for both sets. Further, these data indicate a
bias toward positive values among rating pairs compared to those of ratings in
general, which have an average value of zero by construction\footnote{See
discussion of bias subtraction in {\bf Data} section.}. This latter feature
indicates that items with large numbers of common users tend to be rated higher
by their common users than by their other users since we have already subtracted
the item's average rating thereby avoiding any simple popularity effects.

@@

Having chosen values for $\mu_s$, $\sigma_s$, $\mu_r$, and $\sigma_r$ we are now
able to estimate the similarity of item pairs. To do this, we must solve Eq.
(\ref{eqn:estimator}) for $\hat{y}$ given a pair of ratings. This can be easily
done in practice using a root finding algorithm\footnote{To solve Eq.
(\ref{eqn:estimator}) for $\hat{y}$ given ratings $r_A$ and $r_B$, we use the
{\tt brentq} root finding function of the widely available Python module {\tt
scipy.optimize}.}. The implied predictions of individual users can then be
aggregated to form a new estimate of the similarity of items $A$ and $B$ as
\begin{align}\label{eqn:userPredSim}
UserPredSim(A,B) = \frac{\sum_{u \in U_{AB}}{p_u\cdot\hat{y}(r_{u,A}, r_{u,B})}}
{\sum_{u \in U_{AB}}{p_u}},
\end{align}
where, as before, $U_{AB}$ is the set of common users for items $A$ and $B$ and
$p_u$ is the {\em user predictivity} defined in Eq. (\ref{eqn:predictivity}). We
leverage the training set to find $p_u$ for each user $u$. This is done by
recording the implied prediction $\hat{y}(r_{u,A},r_{u,B})$ for each common user
$u$ for each item pair $(A, B)$ in the training set and then computing each
user's average absolute error $\epsilon_u$ as per Eq. (\ref{eqn:userError}).
Finally, we apply supervised learning techniques to optimize the value of the
parameter $\lambda$ that appears in the definition of $p_u$. This is achieved by
minimizing the error in Eq. {\ref{eqn:userPredSim} made against the training
set. We find that $\lambda = 0.01$, which yields a strong preference for high
{\em user predictivity}, works well for the data studied here.

Figure \ref{fig:MSE_SimR} shows the results of our experiment using
$UserPredSim$ to estimate $TrueSimRand$. To be clear, these plots are showing
the mean squared error (MSE) given by
\begin{align}
\mathrm{MSE} = \frac{1}{\left|S\right|}\sum_{(A,B) \in S}
{\left(Estimator(A,B) - GoldSim(A,B)\right)^2},
\end{align}
where $S$ is the test set of item pairs and $Estimator(A, B)$ represents an
arbitrary estimator of the similarity between items $A$ and $B$. In the upper
panel, we show the MSE for the $N \geq 40$ data. The black dashed line
represents the estimates of $PearSim$. For example, with $n = 20$ common users,
the MSE of $PearSim$ is about $0.07$, which corresponds to an average absolute
error of $0.26$ in the similarity score---roughly the magnitude of $\sigma_s/2$
(see Fig. \ref{fig:Histograms}). These data clearly show that the reliability of
$PearSim$ in estimating $TrueSimRand$ deteriorates rapidly as $n$ decreases.
Below about $n = 20$, the horizontal red line, which represents the MSE obtained
from always guessing the average $GoldSimRand$ value from the training set,
outperforms $PearSim$. These two estimators serve as simple benchmarks.
